---
title: "Halloween Candy Mini-Project"
author: "Yujia Liu (PID: A16967405)"
format: gfm
---

### 1. Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

###2. What is your favorate candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favourite candy: One dime

```{r}
candy["One dime", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

###Using the `skim()` function:

```{r}
library("skimr")
skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Variable winpercent, it is on 0-100 scale while other variables are all between 0-1.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

Zero for candy$chocolate column means the candy does not contain chocolate, while one means the candy contain chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
```

```{r}
ggplot(candy,
       aes(x = candy$winpercent)) +
  geom_histogram()
```

> Q9. Is the distribution of winpercent values symmetrical?

Not symmetrical, the distribution is skewed to the right.

> Q10. Is the center of the distribution above or below 50%?

The center of the distribution is below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc_candy <- candy$winpercent[as.logical(candy$chocolate)]
fruit_candy <- candy$winpercent[as.logical(candy$fruity)]

mean_winpercent_chocolate <- mean(choc_candy)
mean_winpercent_fruity <- mean(fruit_candy)

mean_winpercent_chocolate
mean_winpercent_fruity

mean_winpercent_chocolate > mean_winpercent_fruity
```
On average chololate containing candy is higher ranked than fruity candy.


> Q12. Is this difference statistically significant?

```{r}
t.test(choc_candy, fruit_candy)
```

###3. Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
library("dplyr")
```

```{r}
candy %>%
  arrange(winpercent) %>%
  head(n = 5)
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>%
  arrange(desc(winpercent)) %>%
  head(n = 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy,
       aes(x = winpercent, y = rownames(candy))) + 
      geom_col() + 
  theme(axis.text.y = element_text(size = 3))
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy,
       aes(x = winpercent, y = reorder(rownames(candy),winpercent))) + 
      geom_col() + 
  theme(axis.text.y = element_text(size = 3))
```
Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

Now let’s try our barplot with these colors. Note that we use fill=my_cols for `geom_col()`. Experiment to see what happens if you use col=mycols.

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  theme(axis.text.y = element_text(size = 3))
```
> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst


### 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 100)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord,c(11,12)], n=5)
```
The least popular from the most expensive is Nik L Nip.

Optional
Q21. Make a barplot again with `geom_col()` this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping `geom_col()` for `geom_point()` + `geom_segment()`.

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point() +
  theme(axis.text.y = element_text(size = 3))
  
```

### 5. Exploring the correlation structure

```{r}
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Variables fruity and chocolate are strongly anti-correlated. Variables pluribus and bar are the next most anti-correlated. There are others also have minus values, but are not strongly correlated.

> Q23. Similarly, what two variables are most positively correlated?

Variables winpercent and chocolate, and bar and chocolate are most positively correlated.


### 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = T )
summary(pca)
```

Plot main PCA score plot of PC1 vs PC2.

```{r}
plot(pca$x[,1:2])
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 100)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

Interactive plot to see the labels:

```{r}
#library(plotly)

#ggplotly(p)
```

Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of chocolate and fruity and the similar effects of chocolate and bar (i.e. we already know they are correlated).

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Variables chocolate, fruity, and bar are picked up strongly by PC1. This plot is consistent with the correlation, as chocolate and bar are strongly positively correlated and chocolate and fruity are strongly negatively correlated.
